---
title: "NEEA Energy Analysis"
author: "Wylie Hampson"
date: '2025-02-19'
output: html_document
---

Case Objective:
Please prepare a case analysis in order to answer the questions below.
This should include documentation of your work in this file as well as a presentation that discusses your approach, logic and recommendations to management.
This case is designed to not require more than 2 hours of your time to successfully complete.  We are not looking for a single right answer, but for insights into how you approach analysis and present your insights.

Situation:
Washington State has an Energy Savings goal of being 70% below the 2006 Code average. NEEA's Codes and Standards team is developing a strategy to support this goal and is seeking to understand which scenario(s) the organization should try to influence the market towards.
The questions stated below are to help assess this program potential.
Disclaimer:  data provided within this file are fictitious 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
```

```{r, include=FALSE}
# I want to separate the data into separate tables to help with organization
base_table <- readxl::read_xlsx("data/market_data.xlsx", range = "B5:C14", sheet = "Energy Use") %>% dplyr::rename(year = Code) %>% 
  janitor::clean_names()

# Add correct years to base table to replace NA values
base_table[2:4,1] <- 2006
base_table[6:9, 1] <- 2021

# Read in each table and attach them to the base table and clean them up. I'm also going to convert all tables to kBtu to compare similar units.
energy_use_electric <- readxl::read_xlsx("data/market_data.xlsx", range = "D5:J14", sheet = "Energy Use") %>% 
  janitor::clean_names() %>% #clean column names to make them easier to work with
  dplyr::bind_cols(base_table) %>% # Bind the base table with the electric table.
  dplyr::select(year, base_heating_system, everything()) %>% # Reorder columns.
  dplyr::mutate(across(3:last_col(), ~ .x * 3.412)) # convert kWh to kBtu

energy_use_electric <- energy_use_electric %>% 
  tidyr::pivot_wider(names_from = year, values_from = energy_use_electric %>% select(-year, -base_heating_system) %>% names(), names_sep = "_")

energy_use_gas <- readxl::read_xlsx("data/market_data.xlsx", range = "K5:M14", sheet = "Energy Use") %>% 
  janitor::clean_names() %>% 
  dplyr::bind_cols(base_table) %>% 
  dplyr::select(year, base_heating_system, everything()) %>% 
  dplyr::mutate(across(3:last_col(), ~ .x * 100))  # convert therm to kBtu
  
energy_use_gas <- energy_use_gas %>% 
  tidyr::pivot_wider(names_from = year, values_from = energy_use_gas %>% select(-year, -base_heating_system) %>% names(), names_sep = "_")

energy_use_electric_and_gas <- readxl::read_xlsx("data/market_data.xlsx", range = "N5:T14", sheet = "Energy Use") %>% 
  janitor::clean_names() %>% 
  dplyr::bind_cols(base_table) %>% 
  dplyr::select(year, base_heating_system, everything()) # no need to convert, already in kBtu
  
energy_use_electric_and_gas <- energy_use_electric_and_gas %>% 
  tidyr::pivot_wider(names_from = year, values_from = energy_use_electric_and_gas %>% select(-year, -base_heating_system) %>% names(), names_sep = "_")

weights <- readxl::read_xlsx("data/market_data.xlsx", range = "B3:D9", sheet = "Weights") %>% dplyr::slice(-1) %>% 
  dplyr::rename(weights_2006 = `2006`,
                weights_2021 = `2021`,
                base_heating_system = `HVAC System`) %>% 
  dplyr::mutate(weights_2006 = as.numeric(weights_2006),
                weights_2021 = as.numeric(weights_2021))
  # tidyr::pivot_longer(cols = c(2,3), names_to = "year", values_to = "value") %>% 
  # janitor::clean_names() %>% 
  # dplyr::arrange(year)
```

Q1: Using the data provided in this file, please estimate how close the state is to achieving the stated goal.

```{r}
total_energy_savings_no_gas_backup_heat_pump <- energy_use_electric_and_gas %>% 
  dplyr::filter(base_heating_system != "Gas with Heat Pump Back Up")

new_row <- tibble(base_heating_system = "All Systems", weights_2006 = NA, weights_2021 = NA)

weights <- bind_rows(weights, new_row)

total_energy_savings_no_gas_backup_heat_pump <- total_energy_savings_no_gas_backup_heat_pump %>%
  dplyr::select(base_heating_system, all_2006, all_2021) %>% 
  dplyr::left_join(weights, by = "base_heating_system") %>% 
  dplyr::mutate(weighted_savings_2006 = all_2006 * weights_2006,
                weighted_savings_2021 = all_2021 * weights_2021)

total_energy_savings_no_gas_backup_heat_pump <- total_energy_savings_no_gas_backup_heat_pump %>% 
  dplyr::bind_rows(summarise(total_energy_savings_no_gas_backup_heat_pump, 
                             base_heating_system = "All Systems", 
                             across(-base_heating_system, sum))) %>% 
  dplyr::mutate(energy_saving_percent = ((all_2006 - all_2021) / all_2006) * 100,
                energy_saving_percent_weighted = ((weighted_savings_2006 - weighted_savings_2021) / weighted_savings_2006) * 100)


# total_energy_savings_no_gas_backup_heat_pump <- total_energy_savings %>% 
#   dplyr::filter(base_heating_system != "Gas with Heat Pump Back Up")

# total_energy_savings_no_gas_backup_heat_pump <- total_energy_savings_no_gas_backup_heat_pump %>% 
#   dplyr::bind_rows(summarise(total_energy_savings_no_gas_backup_heat_pump, base_heating_system = "All Systems", across(-base_heating_system, sum)))
```

Q2: Based on that analysis and trends, please identify a scenario you would recommend for the state to meet this goal. What would you present to management for consideration?

```{r}

calculate_total_weighted_saving_percent <- function(gas_furnace = 40196.08,
                                                    gas_furnace_w_ac = 41558.87,
                                                    heat_pump_elec = 35619.49,
                                                    zonal_elec = 35099.94,
                                                    gas_w_pump_backup = 34272.75,
                                                    gas_furnace_weight = 0.08,
                                                    gas_furnace_w_ac_weight = 0.04,
                                                    heat_pump_elec_weight = 0.81,
                                                    zonal_elec_weight= 0.07,
                                                    gas_w_pump_backup_weight = 0) {
  
  
  
}

```
